### 2013-01-20 edupar-13/410.tex
It bloody fits on 8 pages (with garth's feedback)

The course has many learning objectives,
ranging from acquiring detailed factual knowledge about
hardware features
through practicing advanced cognitive processes
such as open-ended design.
%
Students study high-level concepts
such as
\shortversion{least privilege}{protection (least privilege, access control lists vs.\ capabilities)},
file-system internals,
and log-based storage.
%
Emphasis is placed on acquiring information from ``primary sources,''
including both manufacturer-provided hardware documentation
and a non-textbook technical-literature reading assignment.
%
Students begin with a ``blank slate'' rather than a
kernel-source template or an existing operating system,
so they must synthesize design requirements from multiple sources
and must
choose their own module boundaries and inter-module conventions.
%
\shortversion{The}{Due to the foundational nature of kernel code, the}
assignment design and grading encourage students to
think about corner cases, including resource exhaustion,
instead of being satisfied by
\shortversion{}{``the right basic idea''}
implementations that handle only auspicious situations.
%
Finally, most relevant to this work,
students gain substantial experience in
analyzing and writing lock-based multi-threaded code and
thread-synchronization objects.
They practice detecting and documenting deadlock and race conditions,
including both thread/thread concurrency and thread/interrupt concurrency.

@@

\subsubsection{Device Drivers and Game}
The second project is a simple game, such as Hangman,
which runs standalone (without an underlying
operating system).
We provide students with selected C~library routines
(e.g., \x{strcmp()}, \x{memmove()}, \x{printf()});
they write a device-driver library consisting of
console output (to a memory-mapped text-mode video display),
keyboard input (interrupt driven),
and a countdown timer.
The game, which varies by semester, provides students
with the ability to test their drivers.
%
% and the need to write enough code to get into trouble
%
This project and the remaining ones are written in
C with a bit of x86-32 assembly code,
compiled and linked into an ELF executable,
stored into a 1.44-megabyte 3.5-inch floppy-disk image,
and booted via GRUB.
If the floppy-disk image is copied to a real floppy disk
or \shortversion{compact disc,}{embedded into an ``El Torito'' bootable compact disc image,}
it can be booted on standard PC hardware.
However, students most often boot the floppy image
using the Simics system emulator (described below).

\subsubsection{User-space Thread Library}
The third project is implementing a 1:1 thread library for
user-space programs.
The \shortversion{}{library} specification is essentially a stripped-down
version of POSIX Pthreads.
The primary goal of the project is
\shortversion{for students to experiment}{providing students with a laboratory for experimenting}
with concurrency and atomicity.
Students begin by designing mutexes using any
x86-32 atomic instructions they choose
(there is substantial variation across groups).
They then write other thread-synchronization
primitives (condition variables, semaphores,
and reader/writer locks), infrastructure
components (stack allocation/recycling and
a thread registry),
and low-level code to launch and shut down
threads.
A second goal is encouraging students to
consider how ingredients are assembled into
abstractions.
%% Good rhetoric (below). -- bblum
\shortversion{There is a semantic gap}{One example of this is a semantic gap}
between the library-level \x{thr_create()}
call (analogous to \x{pthread_create()})
and the \x{thread_fork} system call provided
by the underlying kernel,
which takes no parameters,
makes no changes to the user's address space,
and cannot meaningfully
be invoked from C~code.
Because these abstractions are so different,
students must
understand each as an independent entity while
deciding how to bridge them.
A similar deliberate semantic gap exists between
\x{cond_wait()} and the system call
which suspends execution of a thread.

Student library code is linked against small
test programs provided by the course staff\shortversion{.}{,
producing one statically linked ELF executable
per test program.}
The test programs are bundled into a RAM-disk
image and linked against a ``reference kernel'' written by
the course staff
and provided in binary form.

Pebbles kernel behavior is specified
via a twelve-page document~\cite{kspec}.
\shortversion{}{The handout for the thread library project itself
is nineteen pages and includes not only a specification
of the thread-library primitives but also
a suggested implementation plan, grading guidelines,
optional design challenges, etc.}
In addition to providing a reliable execution
substrate,
the reference kernel schedules
\shortversion{}{the execution of} user-space threads created by
student code according to a variety of
interleaving policies.
\shortversion{}{In order to aid debugging of concurrency problems,
the \x{misbehave()} system call enables
students to manually select an interleaving policy.}

@@

\subsubsection{Pebbles Extension}
Students who complete the kernel project on time
then work on a smaller kernel-extension project
with various content depending on the semester.
Past projects have included
writing a sound card driver,
a filesystem,
hibernation (suspend to disk),
kernel profiling,
and an in-kernel debugger.
Two recent, more aggressive, projects have been
adding paravirtualization so that their kernels
can host guest kernels and
adding multi-processor support to their single-processor kernels.

In the multi-processor project, students extended their kernels to schedule multiple threads across up to eight cores.
We provided base code for determining a system's CPU count, for booting the application processors (APs), and for sending inter-processor interrupts (IPIs).
% Students needed to design a way to synchronize removing virtual memory mappings in the TLB,
% to implement moving threads between processors,
% and
% to redesign their scheduler's locking primitives to account for the fact that disabling interrupts on a CPU no longer necessarily prevents concurrent accesses.
Students must move threads between processors,
synchronize the removal of memory mappints from multiple TLBs,
and must account for the fact that disabling interrupts on a CPU doesn't prevent concurrent data accesses.
To allow struggling groups to make progress on the project without solving every synchronization challenge, we allowed the use of a ``big kernel lock,'' and rewarded groups according to how little they needed to use it.
Submissions were also graded according to
correctness of the scheduler and spinlocks,
stability on a suite of stress tests,
and the use of IPIs for efficient processor coordination.

@@

As mentioned above, the main execution and debugging
platform used in the class is
Simics~\cite{simics},
originally developed by the Swedish Institute of
Computer Science and
currently a product of Wind River.
Unlike some emulators,
which focus on fast execution of correct code,
Simics
provides very faithful bit-level support
not only for code that behaves correctly
\shortversion
{but also for student kernels that accidentally ``abuse'' hardware.}
{but also for code that is ``incorrect within reason.''
Occasionally student code manages to invoke an unsupported
function of some hardware device or even to cause the
simulator to crash, but this is quite rare, and in almost
every case when the ``bad event'' is characterized it turns
out that what the software requested from the hardware was
a mistake.
Meanwhile, inside its large envelope, Simics supports
\textit{very} detailed models of various x86 processors.}
Unlike hardware virtualization environments,
Simics contains substantial debugger support:
single-stepping,
printing of source-level symbolic
expressions,
stack tracing,
display of TLB entries,
and even summaries of x86 hardware-defined descriptor tables.
Unlike some symbolic debuggers,
Simics is scriptable in Python,
and it is possible to interpose user-written measurement
frameworks to evaluate proposed hardware
%% TODO Fix citations. But also, maybe this mention isn't relevant and
%% could be left out (if we're in space trouble?) -- bblum
features~\shortversion{\cite{SIMFLEX}}{\cite{SIMFLEX,UW-GEMS,FeS2}}.
Performance is very reasonable (Simics can simulate certain
cases, such as a CPU halted awaiting a clock interrupt,
much faster than real time).
%% TODO: One huge advantage of simics over QEMU that we should mention is
%% that QEMU only issues timer interrupts at basic block boundaries. -- bblum

@@

\shortversion{}{Students write code that runs on one specific
hardware platform.}
We target x86 PCs because machines of this type are
available from many manufacturers at many price points.
This means that it is easy for students to obtain access
to a machine (virtual or real) which will let them boot
up their code, type commands, and observe the results.
Also, unlike many System-on-a-Chip platforms,
detailed hardware descriptions are readily available.
% rather than being proprietary or "documented"
% by labyrinthine Linux source code
\shortversion{}{Finally, the platform is fairly stable---with only a
little care, it is possible to write a compliant
kernel which runs on anything newer than a 486DX processor,
i.e., almost any PC hardware which is likely to be in
working condition.}

@@

The course still targets the 32-bit
subset of the x86 platform, for three reasons.
%% TODO I'm not sure I believe this is an advantage. Instead of having
%% to verify if each register is correct, students now have to verify
%% both registers and stack slots. Otherwise this paragraph is good. -- bblum
First, while the small number of registers is
frustrating for compiler writers,
\shortversion{}{from the point of view of OS students,}
the existence of fewer registers means less work
when it is necessary to verify that each register's
value is correct.
Likewise, the calling convention's old-fashioned focus on the stack
makes it easier to track function parameter values.
Second, implementing virtual memory on x86-64 as
opposed to x86-32 requires more code and more
debugging time without imparting greater insight.
\shortversion{}{Finally, since it is impossible for an x86 processor
to be in 64-bit mode with virtual memory off,
students would need to be coached through the
32-bit--to--64-bit transition or would need
to be given a template virtual-memory implementation.}
All in all, at present we are satisifed with
the 32-bit choice.

@@

One cost of programming to the complexities of
real hardware is that
within the scope of a single semester,
students do not have the opportunity to
implement common features of many kernels,
such as a file system, interprocess communication,
a real-time scheduler,
or a network protocol stack.

>>>>

The course has many learning objectives,
ranging from acquiring detailed factual knowledge about
hardware features
through practicing advanced cognitive processes
such as open-ended design.
%
Students study high-level concepts
such as
\shortversion{least privilege}{protection (least privilege, access control lists vs.\ capabilities)},
file-system internals,
and log-based storage.
%
Emphasis is placed on acquiring information from ``primary sources,''
including both manufacturer-provided hardware documentation
and a non-textbook technical-literature reading assignment.
%
Students begin with a ``blank slate'' rather than a
kernel-source template or an existing operating system,
so they must synthesize design requirements from multiple sources
and must
choose their own module boundaries and inter-module conventions.
%
\shortversion{The}{Due to the foundational nature of kernel code, the}
assignment design and grading encourage students to
think about corner cases, including resource exhaustion\shortversion{.}{,
instead of being satisfied by
``the right basic idea''
implementations that handle only auspicious situations.}
%
Finally, most relevant to this work,
students gain substantial experience in
analyzing and writing lock-based multi-threaded code and
thread-synchronization objects.
They practice detecting and documenting deadlock and race conditions,
including both thread/thread concurrency and thread/interrupt concurrency.

@@

\subsubsection{Device Drivers and Game}
The second project is a simple game, such as Hangman,
which runs without an underlying
operating system.
We provide students with selected C~library routines
(e.g., \x{strcmp()}, \x{memmove()}, \x{printf()});
they write a device-driver library consisting of
console output (to a memory-mapped text-mode video display),
keyboard input (interrupt driven),
and a countdown timer.
\shortversion{The game enables students to test their drivers.}
{The game, which varies by semester, provides students
with the ability to test their drivers.}
%
% and the need to write enough code to get into trouble
%
This project and the remaining ones are written in
C with a bit of x86-32 assembly code,
compiled and linked into an ELF executable,
stored into a 1.44-megabyte 3.5-inch floppy-disk image,
and booted via GRUB.
If the \shortversion{}{floppy-disk} image is copied to a real floppy
or \shortversion{compact disc,}{embedded into an ``El Torito'' bootable compact disc image,}
it can be booted on standard PC hardware.
However, students most often \shortversion{use}{boot the floppy image
using} the Simics system emulator (described below).

\subsubsection{User-space Thread Library}
The third project is implementing a 1:1 thread library for
user-space programs,
essentially a stripped-down
version of POSIX Pthreads.
The primary goal is
\shortversion{for students to experiment}{providing students with a laboratory for experimenting}
with concurrency and atomicity.
Students begin by designing mutexes using any
x86-32 atomic instructions they choose
(there is substantial variation across groups).
They then write other thread-synchronization
primitives (condition variables, semaphores,
and reader/writer locks), infrastructure
components (stack allocation/recycling and
a thread registry),
and low-level code to launch and shut down
threads.
A second goal is encouraging students to
consider how ingredients are assembled into
abstractions.
%% Good rhetoric (below). -- bblum
\shortversion{There is a semantic gap}{One example of this is a semantic gap}
between the library-level \x{thr_create()}
call (analogous to \x{pthread_create()})
and the \x{thread_fork} system call provided
by the underlying kernel,
which takes no parameters,
makes no changes to the user's address space,
and cannot meaningfully
be invoked from C~code.
Because these abstractions are so different,
students must
understand each as an independent entity while
deciding how to bridge them.
A similar deliberate semantic gap exists between
\x{cond_wait()} and the system call
which suspends execution of a thread.

Student library code is linked against small
test programs provided by the course staff\shortversion{.}{,
producing one statically linked ELF executable
per test program.}
The test programs \shortversion{run on}{are bundled into a RAM-disk
image and linked against} a ``reference kernel'' written by
the course staff
and provided in binary form.
%
Pebbles kernel behavior is specified
via a twelve-page document~\cite{kspec}.
\shortversion{}{The handout for the thread library project itself
is nineteen pages and includes not only a specification
of the thread-library primitives but also
a suggested implementation plan, grading guidelines,
optional design challenges, etc.}
In addition to providing a reliable execution
substrate,
the reference kernel schedules
\shortversion{}{the execution of} user-space threads created by
student code according to a variety of
interleaving policies.
\shortversion{}{In order to aid debugging of concurrency problems,
the \x{misbehave()} system call enables
students to manually select an interleaving policy.}

@@

\subsubsection{Pebbles Extension}
Students who complete the kernel project on time
then work on a kernel-extension project\shortversion{.}
{with various content depending on the semester.}
Past projects have included
writing a sound card driver,
a filesystem,
hibernation (suspend to disk),
kernel profiling,
and an in-kernel debugger.
Two recent, more aggressive, projects have been
adding paravirtualization so that their kernels
can host guest kernels and
adding multi-processor support to their single-processor kernels.

In the multi-processor project, students extended their kernels to schedule multiple threads across up to eight cores.
We provided base code for determining a system's CPU count, for booting the application processors (APs), and for sending inter-processor interrupts (IPIs).
% Students needed to design a way to synchronize removing virtual memory mappings in the TLB,
% to implement moving threads between processors,
% and
% to redesign their scheduler's locking primitives to account for the fact that disabling interrupts on a CPU no longer necessarily prevents concurrent accesses.
Students must move threads between processors,
synchronize the removal of memory mappings from multiple TLBs,
and account for the fact that disabling interrupts on a CPU doesn't prevent concurrent data accesses.
To allow struggling groups to make progress without solving every synchronization challenge, we allowed the use of a ``big kernel lock,'' and rewarded groups according to how little they needed to use it.
Submissions were also graded according to
correctness of the scheduler and spinlocks,
stability on a suite of stress tests,
and the use of IPIs for efficient processor coordination.

@@

\shortversion{The}{As mentioned above, the} main execution and debugging
platform used in the class is
Simics~\cite{simics},
originally developed by the Swedish Institute of
Computer Science and
currently a product of Wind River.
Unlike some emulators,
which focus on fast execution of correct code,
Simics
provides very faithful bit-level support
not only for code that behaves correctly
\shortversion
{but also for kernels that accidentally ``abuse'' hardware.}
{but also for code that is ``incorrect within reason.''
Occasionally student code manages to invoke an unsupported
function of some hardware device or even to cause the
simulator to crash, but this is quite rare, and in almost
every case when the ``bad event'' is characterized it turns
out that what the software requested from the hardware was
a mistake.
Meanwhile, inside its large envelope, Simics supports
\textit{very} detailed models of various x86 processors.}
Unlike hardware virtualization environments,
Simics contains substantial debugger support:
single-stepping,
printing of source-level symbolic
expressions,
stack tracing,
display of TLB entries,
and even summaries of x86 hardware-defined descriptor tables.
Unlike some symbolic debuggers,
Simics is scriptable in Python,
and it is possible to interpose user-written measurement
frameworks to evaluate proposed hardware
%% TODO Fix citations. But also, maybe this mention isn't relevant and
%% could be left out (if we're in space trouble?) -- bblum
features~\shortversion{\cite{SIMFLEX}}{\cite{SIMFLEX,UW-GEMS,FeS2}}.
\shortversion{}{Performance is very reasonable (Simics can simulate certain
cases, such as a CPU halted awaiting a clock interrupt,
much faster than real time).}
%% TODO: One huge advantage of simics over QEMU that we should mention is
%% that QEMU only issues timer interrupts at basic block boundaries. -- bblum

@@

\shortversion{}{Students write code that runs on one specific
hardware platform.}
We target x86 PCs because machines of this type are
available from many manufacturers at many price points.
This means that it is easy for students to obtain access
to a machine (virtual or real) which will let them boot
up their code, type commands, and observe the results.
\violence{}{Also, unlike many System-on-a-Chip platforms,
detailed hardware descriptions are readily available.}
% rather than being proprietary or "documented"
% by labyrinthine Linux source code
\shortversion{}{Finally, the platform is fairly stable---with only a
little care, it is possible to write a compliant
kernel which runs on anything newer than a 486DX processor,
i.e., almost any PC hardware which is likely to be in
working condition.}

@@

The course still targets the 32-bit
subset of the x86 platform, for multiple reasons.
%% TODO I'm not sure I believe this is an advantage. Instead of having
%% to verify if each register is correct, students now have to verify
%% both registers and stack slots. Otherwise this paragraph is good. -- bblum
First, while the small number of registers is
frustrating for compiler writers,
\shortversion{}{from the point of view of OS students,}
the existence of fewer registers means less work
when it is necessary to verify that each register's
value is correct.
Likewise, the calling convention's old-fashioned focus on the stack
makes it easier to track function parameter values.
Second, implementing virtual memory on x86-64 as
opposed to x86-32 requires more code and more
debugging time without imparting greater insight.
\shortversion{}{Finally, since it is impossible for an x86 processor
to be in 64-bit mode with virtual memory off,
students would need to be coached through the
32-bit--to--64-bit transition or would need
to be given a template virtual-memory implementation.}
All in all, at present we are satisifed with
the 32-bit choice.

@@

One cost of programming to the complexities of
real hardware is that
within the scope of a single semester,
students do not \violence{}{have the opportunity to}
implement common features of many kernels,
such as a file system, interprocess communication,
a real-time scheduler,
or a network protocol stack.
